Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-12-14 Scripts/Manuscript.Rmd
Thomas and Alessandro comments

<!-- # Instructions  -->

<!-- Research articles should present a major advance and can be short or long (maximum 15,000 words) format. Submissions must include an abstract (up to 150 words), an introduction, and sections with brief informative subheadings. You may include up to ten figures and/or tables and about 60 references. Supplementary materials should be limited to information that is not essential for the general understanding of the research presented in the main text and can include data sets, figures, tables, videos, or audio files.  You may include up to a total of ten figures and/or tables (combined) throughout the supplemental text.  -->

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr"#,
 # "kableExtra"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

<!-- * corresponding author: j.palacios@oceans.ubc.ca -->

<!-- # Abstract (150  words) -->

<!-- An opening sentence that states the question/problem addressed by the research AND Enough background content to give context to the study AND A brief statement of primary results AND A short concluding sentence. -->

<!-- Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.  -->

@@

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming, acidification and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]). As a result of climate change, many marine species have changed their distributions towards higher latitude, deeper water or following local temperature gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt], impacting fisheries production and compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds].The projected risks of impacts can be reduced through improving the effectiveness of current fisheries managemen [@Gaines:2018sg; @Cheung:2018dh], including the management of fisheries on species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization (FAO), shared stocks can be classified into three non-exclusive categories: (*i*) transboundary stocks, those that cross neighboring EEZs; (*ii*) straddling stocks, that, in addition to neighboring EEZs, also visit the adjacent high seas; and (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs [@Munro:2004th]. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species, yielding an annual average of 48.5 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 (Chapter 2).

The effectiveness of managing fisheries targeting shared stocks is challenged by the shifts in marine stocks distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quota on shared stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change in the fish stocks and fisheries [@Sumby:2021gw, Chapter 2]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by worsening climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges posed on shared stocks fisheries management will continue to increase in the $21^{st}$ century. To prepare ocean governance to meet the challenges in resource management arising from shifting shared stocks, it is important to understand the time when climate change will affect the sharing distribution of these stocks and the intensity of the resulting impacts [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitatively examined through the concepts of ‘time of emergence’ (ToE) and ‘threat point’ (TP). Time of emergence is defined as the moment in time when an anthropogenic change signal (e.g., future trend) rises above the background noise of natural variability (e.g., historical variation) [@IPCC:2019tn]. This concept has been applied to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind the time of emergence is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background noise of natural climate variability [@Hawkins:2012hc]. The concept of threat point comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage shared stocks [@Sumaila:2013hv]. Specifically, Sumaila et al. (*in press*) applied this concept to understand the impacts of climate change to the economics of shared species between Canada and the United States.

Here, I estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. I employ a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected species shared (Chapter 2) by 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP8.5 (see Methods). I treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population (Chapter 2) and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks, excluding the high seas in this study). Our results suggest that many transboundary stocks have already ‘emerged’ from historical natural variability and overpassed their share threat point in most neighboring EEZs, posing a threat to the sustainability of shared stocks and the resilience of international treaties.

@@

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 (Chapter 2). Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd, Chapter  1], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (Gadus morhua) as an example, based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of G. morhua, but the United States and Greenland do not share a stock of this transboundary species. I defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; \@ref(fig:toeFigureS1))) including island territories. I determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and by natural biome [@Reygondeau:2019uh]. The habitat association of each species was determined based on information available from FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, I used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. I report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

I projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g., sea surface temperature) whereas demersal species distribution is driven by ocean bottom variables (e.g., bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (see Cheung et al. [-@Cheung:2010dt; -@Cheung:2016jd] for model details). DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd], Cheung et al. *in revision*].

The DBEM was forced with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the ‘high’ climate change emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. In the GFDL-ESM2M, atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this chapter relies on understanding the spatial and temporal variation of a stocks’ distribution, we have to understand both the historical as well as the future climate-driven distribution variability, to infer differences between time frames. I do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally small differences in Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

\begin{equation}
\begin{split}
\label{eq:four}
EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = t when Emergence index[t] > 1 s.d.
\end{split}
\end{equation}

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., "threat point" [@Sumaila:2013hv]. Here, I define threat point as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@Sumaila:2020tp, Chapter Four]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

First I estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. I did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ (Chapter 2). Second, I average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($SSR_{th}$) spans from 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. I then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). I chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, I defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, I estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

@@

The time of emergence results were tested for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, I tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1)). In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g. Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure Figure \@ref(fig:toeFigure1)). I based our transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g. the top 90%) of the stock (Figure \@ref(fig:toeFigureS3).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone (EEZ) and transboundary stock: A) results in absolute numbers (green). B) results expressed as a percentage of the total emerging stocks and EEZ with emerging stock (purple)."}

@@

Our analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary stocks between 2005 and 2010 will see their distribution range shifts beyond historical variabilities between 2025 and 2040 (Figure \@ref(fig:toeFigure3)). In some cases, like Indonesia (ranked $8^{th}$ in terms of revenue from transboundary stocks), Thailand (ranked $7^{th}$ in terms of revenue from transboundary stocks) and the United States (ranked $4^{th}$ in terms of fishing revenue from transboundary stocks), stocks’ range shifts swill emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more stocks with range shifts emerging from historical variabilities, being Spain (n = 122), France (n = 85), China and Portugal (both n = 73) and Senegal (n = 66) the top 5 leading countries in number of emerging stocks.

>>>>

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr"#,
)

@@

@@

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming, acidification and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]). As a result of climate change, many marine species have changed their distributions towards higher latitude, deeper water or following local temperature gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt], impacting fisheries production and compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries management [@Gaines:2018sg; @Cheung:2018dh], including the fisheries management of species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization (FAO), shared stocks can be classified into four non-exclusive categories: (i) transboundary stocks, those that cross neighboring EEZs; (ii) straddling stocks, that, in addition to neighboring EEZs, also visit the adjacent high seas; (iii) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and finally (iv) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species representing 67% or the identified fished taxa, yielding an annual average of 48.5 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary stocks is challenged by the shifts in marine stocks distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quotas for transboundary stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change on the fish stocks and fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by intensifying climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges for the fisheries management of transboundary stocks will continue to increase during the 21st century. In order to prepare ocean governance to meet the resource management challenges arising from shifting transboundary stocks, it is important to constrain when climate change will affect the dynamics of transboundary  stocks and the intensity of the resulting impacts [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitatively examined through the concepts of ‘time of emergence’ (ToE) and ‘threat point’ (TP). Time of emergence is defined as the moment in time when a signal (e.g., future anthropogenic trend) emerges from the background noise of natural variability (e.g., historical natural variation) [@Hawkins:2012hc] and has been applied to multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind the time of emergence is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background noise of natural climate variability [@Hawkins:2012hc]. The concept of threat point comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage transboundary stocks [@Sumaila:2013hv], including the impacts of climate change on the economics of transboundary species between Canada and the United States (Sumaila et al., *in press*) .

Here, we estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. We employ a species distribution model driven by output from a comprehensive Earth System Model with ten ensemble members to project the distribution of 663 transboundary species [@PalaciosAbrantes:2020tv] across 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP8.5 (see Methods). We treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks that exclude  the high seas. We find that many transboundary stocks have already ‘emerged’ from historical natural variability and exceeded their share threat point in most neighboring EEZs, posing a threat to their sustainability and the resilience of international treaties.

@@

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 (Chapter 2). Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd, Chapter 2], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (Gadus morhua) as an example, and based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of Atlantic cod, but the United States and Greenland do not share a stock of this transboundary species. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Figure \@ref(fig:toeFigureS1) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome [@Reygondeau:2019uh]. The habitat association of each species was determined following the classification in FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences for sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface whereas demersal species distribution is driven by ocean bottom variables . DBEM then estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid (see Cheung et al. [-@Cheung:2010dt; -@Cheung:2016jd] for model details). Importantly, DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd, Cheung et al. *in revision*].

The DBEM was forced with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the high greenhouse gas emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. Because the main approach of this paper relies on understanding the spatial and temporal variation of a stocks’ distribution, we have to understand distribution variability during both the historical and the future periods, to infer differences between time frames. I do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for my experiment, each of the ten GFDL-ESM2M ensembles were started from infinitesimally small differences in Earth system initial conditions in year 1950 resulting in a unique atmosphere and ocean state at each point in time after about three years for surface waters and eight years for subsurface waters [@Frolicher:2020cc]. By design, variations among ensemble members are then solely due to natural variability. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced anthropogenic climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

\begin{equation}
\begin{split}
\label{eq:four}
EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = ToE_{t} > 1 s.d.
\end{split}
\end{equation}

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than those operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., “threat point” [@Sumaila:2013hv]. Here, we define the threat point as the minimum required proportion of a shared stock present within an EEZ for a country to engage in cooperative management with their sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a unilateral management of the shared stock.

First, we estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. We did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ (Chapter 2). Second, we averaged the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of variability. The first time period ($SSR_{th}$) spans 1951 to 2005 and is chosen to match the historical period in the GFDL ESM2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed to represent the historic baseline of shared distributions for each stock. I then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). I chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, I defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, I estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

@@

The time of emergence results were tested for normality (e.g. skewness, kurtosis) and performed two non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, I tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1)). This means that on average, the distribution of transboundary stocks will diverge from their natural internal variability by 2029. In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g., Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slows from a 1% to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure \@ref(fig:toeFigure1)). I based the transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that this method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g., the top 95%) of the stock (Figure \@ref(fig:toeFigureS3).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence of the transboundary index by Exclusive Economic Zone (EEZ) and transboundary stock"}

@@

My analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary stocks between 2005 and 2010 will see the distribution range of their stocks shift beyond historical variabilities between 2025 and 2040 (Figure \@ref(fig:toeFigure3)). In some cases, like Indonesia (ranked $8^{th}$ in terms of revenue from transboundary stocks), Thailand (ranked $7^{th}$ in terms of revenue from transboundary stocks) and the United States (ranked $4^{th}$ in terms of fishing revenue from transboundary stocks), the range shift of transboundary stocks will emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more transboundary stocks with range shifts emerging from historical variability, such as Spain (n = 122), France (n = 85), China and Portugal (both n = 73) and Senegal (n = 66), which are the top 5 leading countries in terms of number of transboundary stocks emerging from historical variability.

On average, stocks that see their transboundary index emerge by 2100 represent 27% $\pm$ 23% of their EEZ’s fisheries revenue that targets transboundary species between 2004 and 2010. However, large variation exists in the contribution of stocks with emerging transboundary index to the current revenue participation across EEZs. In some cases, e.g., El Salvador and New Zealand, stock’s with emerging transboundary index represent less than 1% of the revenue generated from transboundary fisheries in their respective EEZs between 2004 and 2014. On the contrary, stocks with emerging transboundary index from the Marshall Islands and Tokelau represent over 90% of the revenue generated from fisheries targeting transboundary stocks within their respective EEZs during the same time period (Figure \@ref(fig:toeFigure2)A). In addition, some EEZs are projected to have a few emerging stocks that represent a large proportion of revenue from transboundary stocks e.g., in Peru and Finland (with 10 and 5 emerging stocks representing 86% and 78% of average fishing revenue between 2004 to 2010, respectively). Overall, Morocco and Malaysia are the only two countries amongst the top 10 countries with the larger number of stocks with emerging transboundary index (n = 56 and 53, respectively) where the contribution of such stocks to fishing revenue is less than 20%. This indicates that that these countries despite having a large proportion of transboundary stocks emerging, the economic contribution of these are relatively small.
